---
title: "Project 2"
author: "Kylin Sun"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
library(readr)
library(dplyr)
library(lubridate)
library(tidyr)
library(ggplot2)
```

## GitHub Repository  
[https://github.com/kylin081/Project-2](https://github.com/kylin081/Project-2)

```{r}
# Read data
aq_2018 <- read_csv("data-6/data/AQ_2018.csv")
aq_2019 <- read_csv("data-6/data/AQ_2019.csv")
aq_2020 <- read_csv("data-6/data/AQ_2020.csv")
```


```{r}
# Data cleaning function
clean_air <- function(df) {
  df %>%                                               # Rename variables
    rename(
      date = Date,
      co   = `Daily Max 8-hour CO Concentration`,
      pm25 = `Daily Mean PM2.5 Concentration`,
      o3   = `Daily Max 8-hour Ozone Concentration`
    ) %>%
    mutate(
      date = mdy(date)                                 # convert the date format
    ) %>%
    group_by(date) %>%                                 # average multiple measurements per day
    summarise(
      co   = mean(co, na.rm = TRUE),
      pm25 = mean(pm25, na.rm = TRUE),
      o3   = mean(o3, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      across(c(co, pm25, o3), ~ ifelse(.x < 0, 0, .x)) # replace negative values with zero
    ) %>%
    distinct()                                         # remove duplicates
}
```


```{r}
# Cleaning thedata for all three years
aq18 <- clean_air(aq_2018) %>% mutate(year = 2018)
aq19 <- clean_air(aq_2019) %>% mutate(year = 2019)
aq20 <- clean_air(aq_2020) %>% mutate(year = 2020)

# Merge datasets together
aq_all <- bind_rows(aq18, aq19, aq20) %>%
  mutate(
    month = month(date),
    month_lab = month(date, label = TRUE, abbr = TRUE)
  )
```

## Plot 1: Monthly Average CO and O3 (2018–2020)

```{r plot1, echo=FALSE}
# Convert CO and O3 into long format for easier summarizing
co_o3_summary <- aq_all %>%
  select(month_lab, co, o3) %>%              # keep relevant variables
  pivot_longer(
    cols = c(co, o3),                         # reshape CO and O3 into one column
    names_to = "pollutant",                   # new column indicating pollutant type
    values_to = "value"                       # new column holding concentrations
  ) %>%
  group_by(month_lab, pollutant) %>%          # group by month and pollutant
  summarise(
    mean = mean(value, na.rm = TRUE),         # monthly average concentration
    sd   = sd(value, na.rm = TRUE),           # standard deviation
    n    = sum(!is.na(value)),                # number of observations
    se   = sd / sqrt(n),                      # standard error
    lower = mean - 1.96 * se,                 # lower bound of 95% CI
    upper = mean + 1.96 * se,                 # upper bound of 95% CI
    .groups = "drop"
  )

# Plot monthly averages with confidence intervals
ggplot(co_o3_summary,
       aes(x = month_lab, y = mean, color = pollutant, group = pollutant)) +
  geom_point(size = 2) +                      # mean points
  geom_line() +                               # connect the points
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) + # CI bars
  labs(
    x = "Month",
    y = "Average Concentration",
    color = "Pollutant",
    title = "Monthly Average CO and O3 (2018–2020)",
    subtitle = "Means with 95% confidence intervals"
  ) +
  theme_minimal()                             # clean theme
```

Plot 1 shows the monthly average concentrations of CO and O3 across all three years combined.
O3 displays a clear seasonal pattern, with higher levels in the summer months and lower levels during winter.
CO shows the opposite pattern, with higher concentrations in winter and lower concentrations in summer.
The 95% confidence intervals are relatively narrow, suggesting that the monthly averages are estimated with good precision.

## Plot 2: Monthly Average PM2.5 by Year

```{r plot2, echo=FALSE}
# Compute monthly averages of PM2.5 for each year
pm25_summary <- aq_all %>%
  group_by(year, month_lab) %>%               # group by year and month
  summarise(
    mean_pm25 = mean(pm25, na.rm = TRUE),     # monthly mean
    .groups = "drop"
  )

# Plot PM2.5 trends for each year
ggplot(pm25_summary,
       aes(x = month_lab, y = mean_pm25, color = factor(year), group = year)) +
  geom_line() +                               # lines connecting months
  geom_point(size = 2) +                       # points for each month
  labs(
    x = "Month",
    y = "Average PM2.5 Concentration",
    color = "Year",
    title = "Monthly PM2.5 Concentrations by Year",
    subtitle = "Comparing seasonal trends across 2018, 2019, and 2020"
  ) +
  theme_minimal()
```

Plot 2 compares the monthly average PM2.5 concentrations for 2018, 2019, and 2020.
The monthly patterns for 2018 and 2019 are relatively similar, although 2019 appears slightly lower in several months.
The 2020 values only cover January through April, so the line is shorter and cannot be directly compared across all months.
Overall, PM2.5 shows noticeable month-to-month variability, but no consistent seasonal pattern across all three years.